---
title: "Exploring the BRFSS data"
author: "Janish Parikh"
date: "30/08/2020"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data
```{r}
brfss2013<-load('x.Rdata')
```

* * *

## Part 1: Data

The behavioral Risk Factor Surveillance System (BRFSS) is a collaborative project between all of the states in the United States( US) and participating US territories and the Centers for Disease Control and Prevention (CDC). It's major objective is to collect uniform, state-specific data on preventive health practices and risk behaviors that are linked to chronic diseases, injuries, and preventable infectious diseases that affect the adult population.
        Now we will dive into some important aspects of our data.

1) **What is our data about**?

 Answer: Our data is a complete survey conducted by BRFSS regarding the health status of the adult US citizens and their health related issues. This data "brfss2013" provides the data of *491775* patients and total of *330* variables are used as a metrics for evaluation. Actually, the whole population was taken into account, however, this data provides information for the US adult citizens. Hence, a sample is taken.

2) **Why is this data collected**?

 Answer: This data is helpful for BRFSS for studying and analyzing several factors that are responsible for ill-health of adult US citizens and to draw the meaningful insights from these data.
 
3) **Mode of data collection**:

  This data is conducted through survey. The survey was taken using landline telephone and cellular telephone.
    
  a) *During landline telephone survey*, interviewers collected data from a randomly selected adult in a household. Disproportionate stratified sampling (DSS) has been used for the landline sample.
    
  b) *During cellular telephone survey*, interviewers collected data from an adult who participates by using a cellular telephone and resided in a private residence or college housing. The sample is randomly generated from a sampling frame of confirmed cellular area.
    
### Scope of inference

While talking about scope of inference, the BRFSS study is of observational nature which employs stratified design based on random sampling.
Also, the survey must be generic to all non-institutionalized adults (18 or more). This whole survey is a complete experimental study.

* * *

* * *
## Part 2: Research questions

**Research question 1:**

One of the first research question involves how sleeping habit of an US citizen have any effects on chronic health condition specifically depressive disorder, specifically for those category of people who sleep for less than 5 hours.

**Research question 2:**

My next research question involves seeing the relativity of the US citizens who have smoked at least 100 cigarettes, had any asthma related issues, and had other types of cancers too.


**Research question 3:**

Third research question is related to mental support based on gender. There are many issues related to mentality in US that we get to hear via News. So, I want to know if male get higher mental support and care than female or vice-verse or maybe equal.

* * *

## Part 3: Exploratory data analysis

**Research quesion 1:**

```{r, message=FALSE}
#Selecting attributes of our interest i.e sleptim1(no of sleeping hours), addepev2(Yes/No) and removing the missing values
sleep_depression <- brfss2013 %>%
  filter(!is.na(sleptim1),!is.na(addepev2)) %>%
  select(sleptim1,addepev2)
  
up_depression<- sleep_depression %>%
  filter(sleptim1<5) %>%
  group_by(addepev2) %>%
  summarise(count=n())

up_depression

ggplot(data=up_depression,aes(x=addepev2,y=count))+
      geom_bar(stat="identity",color='black',fill='sky blue')+
      xlab("Depressive Order for people having <5 hours of sleep?")+
      ylab("Number of US citizens ")
```

**From the bar diagram**, we can see that people not having depressive order even after having less than five hours of sleep is more than people having depressive order under same condition which is I found, quite, ***astounding***. This can visualise us an image that sleeping disorder has less influence on depression. But again, we cannot explicitly say that since it can be influenced by many other health factors.



**Research Question 2:**


```{r , message=FALSE}
sm_as_cr <- brfss2013 %>%
  filter(!is.na(smoke100),!is.na(asthma3),!is.na(chcocncr)) %>%
  select(smoke100,asthma3,chcocncr)

sm_as_cr<- sm_as_cr %>%
  mutate(allyes= ifelse(smoke100=="Yes"& asthma3=="Yes" & chcocncr=="Yes","Yes","No"))

all_yes<-sm_as_cr %>%
  group_by(allyes) %>%
  summarise(count=n())

ggplot(data=all_yes,aes(x=allyes,y=count))+
      geom_bar(stat='identity', color ="black", fill="turquoise")+
      xlab("Smoking+Asthma+Cancer")+
      ylab("Number of US Citizens")
```


From the graph, we can say that those people who had smoked at least 100 times, had asthama too, and cancer related complications **are very less** compared to those that do not have all these. From here, we can estimate that there are very, very less people who are bounded with all these critical health problems. From this, we can also guess that asthma and cancer are not really caused by smoking. But again, we can't explicitly say anything since the sample can be influenced by many factors such as habitat, financial status and many more.


**Research question 3:**

```{r, message=FALSE}
life<- brfss2013 %>%
  filter(!is.na(lsatisfy),!is.na(emtsuprt),!is.na(cadult)) %>%
  select(lsatisfy,emtsuprt,cadult)

life <- life %>%
  mutate(male= ifelse(cadult=="Yes, male respondent",1,0))

gender_supp<- life %>%
  filter(lsatisfy=="Very satisfied" & emtsuprt=="Always") %>%
  group_by(male) %>%
  summarise(count=n())

val=c("Female","Male")

ggplot(data=gender_supp,aes(x=male,y=count,fill=val))+
      geom_bar(stat='identity')+
      xlab("Gender")+
      ylab("Emotional Support")
```

**From the bar graph**, we can clearly see that the number of males and females are approximately equal. From this visualization, we can , to a certain extent say that, male and female in US states get equal emotional support. However, this is only acceptable form for the data collected by BRFSS. It cannot be generalized to the whole population. This can be influenced by many external factors.

**Research question 4:**
Among non-institutionalized adults in the US, is there any association between respondents’ health status and their consumption of vegetables and fruits. It would be interesting to find out how the health status category is affected by vegetable and fruits intake.

```{r, message=F}
q2_a <- brfss2013 %>% filter(!is.na(genhlth), !is.na(X_vegesum), genhlth == 'Excellent') %>%
  summarise(mean_Excellent_veg = mean(X_vegesum))

q2_b <- brfss2013 %>% filter(!is.na(genhlth), !is.na(X_vegesum), genhlth == 'Very good') %>%
  summarise(mean_VeryGood_veg = mean(X_vegesum))

q2_c <- brfss2013 %>% filter(!is.na(genhlth), !is.na(X_vegesum), genhlth == 'Good') %>%
  summarise(mean_Good_veg = mean(X_vegesum))
q2_d <- brfss2013 %>% filter(!is.na(genhlth), !is.na(X_vegesum), genhlth == 'Fair') %>%
  summarise(mean_Fair_veg = mean(X_vegesum))
q2_e <- brfss2013 %>% filter(!is.na(genhlth), !is.na(X_vegesum), genhlth == 'Poor') %>%
  summarise(mean_Poor_veg = mean(X_vegesum))
q2_f <- brfss2013 %>% filter(!is.na(genhlth), !is.na(X_frutsum), genhlth == 'Excellent') %>%
  summarise(mean_Excellent_fruit = mean(X_frutsum))
q2_g <- brfss2013 %>% filter(!is.na(genhlth), !is.na(X_frutsum), genhlth == 'Very good') %>%
  summarise(mean_VeryGood_fruit = mean(X_frutsum))
q2_h <- brfss2013 %>% filter(!is.na(genhlth), !is.na(X_frutsum), genhlth == 'Good') %>%
  summarise(mean_Good_fruit = mean(X_frutsum))
q2_i <- brfss2013 %>% filter(!is.na(genhlth), !is.na(X_frutsum), genhlth == 'Fair') %>%
  summarise(mean_Fair_fruit = mean(X_frutsum))
q2_j <- brfss2013 %>% filter(!is.na(genhlth), !is.na(X_frutsum), genhlth == 'Poor') %>%
  summarise(mean_Poor_fruit = mean(X_frutsum))
mean_veg_vector_health <- c(q2_a$mean_Excellent_veg, q2_b$mean_VeryGood_veg, q2_c$mean_Good_veg,
                         q2_d$mean_Fair_veg, q2_e$mean_Poor_veg)
names <- c("Excellent", "VeryGood", "Good", "Fair", "Poor")
par(mfrow=c(2,1))
barplot(mean_veg_vector_health, names.arg = names, cex.names = .9, xlab = "Health status category",
        ylab = 'Mean of Veg Consumption', main= 'Mean of Veg Consumption by Health status categories')
lines(mean_veg_vector_health, col = 'Blue')
mean_fruit_vector_health <- c(q2_f$mean_Excellent_fruit, q2_g$mean_VeryGood_fruit, q2_h$mean_Good_fruit,
                           q2_i$mean_Fair_fruit, q2_j$mean_Poor_fruit)
barplot(mean_fruit_vector_health, names.arg = names, cex.names = .9, xlab = "Health status category",
        ylab = 'Mean of Fruits Consumption', main = 'Mean of Fruits Consumption by Health status categories')
lines(mean_fruit_vector_health, col = 'Blue')
```

We can see that “Excellent” health people consume relatively more fruits & vegetables. On the average, “Poor” health people consume less fruits & vegetables. This relation of fruit & vegetable consumption with the health status category is not causal.

**Research Question 5 :**

Among non-institutionalized adults in the US, which state has the highest prevalence of adults drinking alcoholic beverages? In each state and US territory, we compare the percentage of respondents who currently drinking alcoholic beverages.

The variable drnkany5 documents adults who drink alcoholic beverages in the past 30 days: it takes the value “Yes” for a respondent who had at least one drink of alcohol in the past 30 days, and “No” for a respondent who had not drink any alcohol in the past 30 days. Let’s examine the values of drnkany5:
```{r, message=F, warning=F}
q3 <- brfss2013 %>% select(X_state, drnkany5) %>% na.omit()
q3 %>% group_by(drnkany5) %>% summarize(count = n())
q3  <- q3 %>% mutate(drnkany5 = ifelse(drnkany5 == "Yes", 1, 0))
drinkers_by_state <- q3 %>%
  filter((X_state != 0) & (X_state != 80)) %>% # Filter out extraneous X_state values
  group_by(X_state) %>%
  summarize(fraction_current_drinkers = mean(drnkany5, na.rm = TRUE)) %>%
  arrange(desc(fraction_current_drinkers))
drinkers_by_state %>% ggplot(aes(x = reorder(X_state, fraction_current_drinkers), y = fraction_current_drinkers)) +
  geom_col() + coord_flip() + xlab("State") + ylab("% of Adults Currently Drinking")
```

We observed that the top 5 states with the highest drinking rate are Puerto Rico, Utah, Mississippi, Kentucky, and Tennessee (in descending order), and the top 5 states with the lowest drinking rate are Minnesota, Wisconsin, District of Columbia, Vermont, and New Hampshire (in ascending order).